Discrete Quantum Theories and Computing

### 2018-02-27 dissertation.tex
Updates from ShareLaTeX
dissertation.lyx "Chapter 2 Conventional Quantum Theory" Updating...

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[twoside]{iuphd}
\usepackage[T1,T2A]{fontenc}
\usepackage[utf8]{inputenc}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{verbatim}
\usepackage{mathrsfs}
\usepackage{pdfpages}
\usepackage{amsmath}
\usepackage{amssymb}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage[ backend=biber]{biblatex}
\addbibresource{prop.bib}
\ExecuteBibliographyOptions{sorting=none,maxbibnames=10,doi=false,isbn=false,url=false}
\DeclareUnicodeCharacter{0229}{\c{e}}
\DeclareUnicodeCharacter{00A0}{~}
\usepackage{mmap}
\usepackage{comment}

@@

\usepackage[all]{xy}
\usepackage{framed}
\usepackage[braket,qm]{qcircuit}
\usepackage{mathrsfs}
\usepackage{setspace}

\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}

@@

\todo{{Table of symbols?}}

@@

Conventional quantum theory (CQT) is described by the following: 
\begin{enumerate}
\item[(i)] $D=2^{n}$ orthonormal basis vectors of an $n$-qubit state, 
\item[(ii)] the normalized $D$ complex probability amplitude coefficients describing
the contribution of each basis vector, 
\item[(iii)] a set of probability-conserving unitary matrix operators that suffice
to describe all required state transformations of a quantum circuit, 
\item[(iv)] and a measurement framework. 
\end{enumerate}
We remark that there are many things that are assumed in CQT, such
as the absence of zero norm states for non-zero vectors, and the decomposition
of complex amplitudes into a pair of ordinary real numbers. One also
typically assumes the existence of a Hilbert space with an orthonormal
basis, allowing us to write $n$-qubit \textit{pure} states in general
as Hilbert space vectors with an Hermitian inner product: 
\begin{eqnarray}
\ket{\Psi} & = & \sum_{i=0}^{D-1}\alpha_{i}\ket{i}\,.
\end{eqnarray}
Here $\alpha_{i}\in\mathComplex$ are complex probability amplitudes,
$\VVec{\alpha}\in\mathComplex^{D}$, and the $\{\ket{i}\}$ is an
orthonormal basis of states obeying 
\begin{equation}
\braket{i}{k}=\delta_{ik}\,,
\end{equation}
where $\bra{i}$ is called the dual vector of $\ket{i}$.

@@

In Sec.~\ref{CQC1qubitBloch.sec}, we focus on the discrete geometric
issues raised by the properties (i) and (ii) given above for CQT.

(iii)

In Sec.~\ref{sec:fuzzy}, we introduces the important issues of (iv)
and the foundations of quantum probability space and fuzzy measurement.

To facilitate the transition to DQT carried out in later sections,
we concern ourselves first with the properties of the simplest possible
abstract state object in CQT, the single qubit state.

@@

@@

\subsection{The single qubit problem}

A single qubit already provides access to a wealth of geometric information
and context. We write the single qubit state as 
\begin{equation}
\begin{array}{lr}
\ket{\psi_{1}}=\alpha_{0}\ket{0}+\alpha_{1}\ket{1}\  & \ \alpha_{0},\alpha_{1}\in\C\end{array}\ .%%
%%\alpha_{1}{array}
\label{A1.eq}
\end{equation}

A convenience for probability calculations and a necessity for computing
relative state properties is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $e^{i\phi}$, that is, 
\begin{equation}
(\alpha_{0},\alpha_{1})\sim(e^{i\phi}\alpha_{0},e^{i\phi}\alpha_{1})\ .\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration, which we can write
down as follows: let 
\begin{equation}
\begin{array}{cc}
\alpha_{0}=x_{0}+iy_{0}, & \alpha_{1}=x_{1}+iy_{1}\end{array}\ .\label{A4.eq}
\end{equation}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
x_{0}^{\ 2}+y_{0}^{\ 2}+x_{1}^{\ 2}+y_{1}^{\ 2}=1\ .\label{A5.eq}
\end{equation}

There is a family of 6 equivalence classes of quadratic maps that
take the remaining 3 degrees of freedom in \Eq{A5.eq} and reduce
them to 2 degrees of freedom by effectively removing $e^{i\phi}$
(``fibering out by the circle $\Sphere{1}$''). The standard form
of this class of maps (``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\ .\nonumber 
\end{eqnarray}
These transformed coordinates obey 
\begin{equation}
\|\Vec{X}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Figure \ref{hopfFibers.fig}
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point\/} $(\theta,\phi)$ on the surface
$X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

%% (We try to list the sources of all computed figures)%% Sources for hopfFibers.fig:  dqc-geom-2012/00mathFigs/Hopf-Fiber.nb
\begin{figure}[htb]
\centerline{ \includegraphics[width=0.48\columnwidth,bb = 0 0 200 100, draft, type=eps]{../../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/hopfS2.pdf}\hspace{0.03in}
\includegraphics[width=0.48\columnwidth,bb = 0 0 200 100, draft, type=eps]{../../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/hopfFibers.pdf}}
\centerline{\hspace{1in} (a) \hfill{}(b) \hspace{1in}} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration as a family
of circles (the paths of $e^{i\phi}$), each corresponding to a single
point on the sphere in (a). Points in (a) are color coded corresponding
to circles in (b), e.g., one pole contains the red elliptical circle
that would become an infinite-radius circle in a slightly different
projection, and the opposite pole corresponds to the large perfectly
round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{eqnarray}
\hspace*{-0.5cm}\left(x_{0},y_{0},x_{1},y_{1}\right) & = & \left(\cos\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\sin\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\ \right.\nonumber \\
 &  & \left.\cos\frac{\theta-\phi}{2}\sin\frac{\psi}{2},\ \sin\frac{\theta-\phi}{2}\sin\frac{\psi}{2}\right),\label{A8a.eq}
\end{eqnarray}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\pi$, with $0\leq\frac{\theta+\phi}{2}<2\pi$ and $0\leq\frac{\theta-\phi}{2}<2\pi$,
we see that 
\begin{eqnarray}
\left(X,Y,Z\right)=\left(\cos\phi\sin\psi,\,\sin\phi\sin\psi,\,\cos\psi\right)\ .\label{A8b.eq}
\end{eqnarray}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to 
\begin{equation}
\ket{\psi_{1}}=e^{i\phi}\cos\frac{\psi}{2}\ket{0}+\sin\frac{\psi}{2}\ket{1}\ ,\label{A9.eq}
\end{equation}
and an irreducible state can be represented as a point on a sphere,
as shown in Figure \ref{blochSphere.fig}(a).

\begin{figure}[thb]
\begin{centering}
\includegraphics[width=0.47\columnwidth,bb = 0 0 200 100, draft, type=eps]{../../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/blochSphere.pdf}
\hspace{0.01in} \includegraphics[width=0.47\columnwidth,bb = 0 0 200 100, draft, type=eps]{../../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/blochSphereArc.pdf} 
\par\end{centering}
\centerline{\hspace{1in} (a) \hfill{}(b) \hspace{1in}} \caption{(a) The conventional Bloch sphere with a unique state represented
by the point at the red sphere. (b) The geodesic shortest-distance
arc connecting two one-qubit quantum states.}
\label{blochSphere.fig} 
\end{figure}

%% AMR: replaced ', probabilistic, exponentially complex transitions from%% state to state' with the above

@@

For one qubit, we could easily solve the problem of reducing the full
unit-norm space to its irreducible components $\Vec{X}=(X,Y,Z)$ characterizing
the Bloch sphere. We have just argued that essentially the same process
is possible for $n$-qubits: in the abstract argument, we simply identify
the family of coefficients $\{\alpha_{i}\}$ as being the same if
they differ only by an overall phase $e^{i\phi}$. However, in practice
this is not a construction that is easy to realize in a practical
computation. We now outline an explicit algorithm for accomplishing
the reduction to the irreducible $n$-qubit state space $\CP{D-1}$;
this construction will turn out to be useful for the validation of
our discrete results to follow below.

@@

We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $e^{i\phi}$ from the $n$-qubit states. If we take a particular
weighting of the elements of the density matrix $\rho$, we can construct
a \textit{unit vector\/} of real dimension $D^{2}$ with the form:
\begin{equation}
\Vec{X}=(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,\label{nqCoords.eq}
\end{equation}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
This construction gives an explicit embedding of the $(D-1)$-dimensional
complex, or $(2D-2)=(2^{n+1}-2)$-dimensional real, object in a real
space of dimension $D^{2}=2^{2n}$. However, this is somewhat subtle
because the vector is of unit length, so technically the embedding
space is a sphere of dimension $D^{2}-1=2^{2n}-1$ embedded in $\R^{D^{2}}$;
the one-qubit irreducible states could be represented in a 4D embedding,
but the magnitude of every coordinate would be one; furthermore, the
object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two\/} coordinate patches, one covering one pole of
$\Sphere{2}$ with coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & 1+0i\\
\alpha_{1} & = & x_{1}+iy_{1}
\end{array}
\end{eqnarray}
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & x_{0}+iy_{0}\\
\alpha_{1} & = & 1+0i\ .
\end{array}
\end{eqnarray}

We finally see that the irreducible $n$-qubit state space $\CP{D-1}$
is described by $D$ projectively equivalent coordinates, one of which
can always be scaled out to leave $(D-1)$ actual (complex) degrees
of freedom. We must choose, in turn, $D$ different local sets of
complex variables defined by taking the value $\alpha_{k}=1$, with
$k=0,\ldots,D-1$, and allowing the remaining $D-1$ complex (or $2D-2$
real) variables to run free. No single set of coordinates will work,
since the submanifold including $\alpha_{k}=0$ is undefined and another
coordinate system must be chosen to cover that coordinate patch. This
is a standard feature of the topology of non-trivial manifolds such
as $\CP{D-1}$ (see any textbook on geometry \cite{Berger}).

@@

>>>>

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[twoside]{iuphd}
\usepackage[T1,T2A]{fontenc}
\usepackage[utf8]{inputenc}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{verbatim}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{pdfpages}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newlength{\lyxlabelwidth}      % auxiliary length

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage[ backend=biber]{biblatex}
\addbibresource{prop.bib}
\ExecuteBibliographyOptions{sorting=none,maxbibnames=10,doi=false,isbn=false,url=false}
\DeclareUnicodeCharacter{0229}{\c{e}}
\DeclareUnicodeCharacter{00A0}{~}
\usepackage{mmap}
\usepackage{comment}
\usepackage[all]{xy}
\usepackage{framed}
\usepackage[braket,qm]{qcircuit}
\usepackage{mathrsfs}
\usepackage{setspace}

@@

\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Sphere}[1]{\mathbf{S}^{#1}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\rmi}{i}
\newcommand{\rme}{e}

@@

\todo{Table of symbols?}

@@

Conventional quantum theory (CQT) is described by the following:
\begin{enumerate}[label=(\roman{enumi})]
\item \label{enu:-orthonormal-basis}$D$ orthonormal basis vectors for
a Hilbert space of dimension~$D$, 
\item \label{enu:probability-amplitude}the normalized $D$ complex probability
amplitude coefficients describing the contribution of each basis vector, 
\item \label{enu:unitary-matrix}a set of probability-conserving unitary
matrix operators that suffice to describe all required state transformations, 
\item \label{enu:measurement-framework}and a measurement framework. 
\end{enumerate}
We remark that there are many things that are assumed in CQT, such
as the absence of zero norm states for non-zero vectors, and the decomposition
of complex amplitudes into a pair of ordinary real numbers. One also
typically assumes the existence of a $D$-dimensional Hilbert space
with an orthonormal basis, allowing us to write \textit{pure} states
in general as Hilbert space vectors with an Hermitian inner product:
\begin{eqnarray}
\ket{\Psi} & = & \sum_{i=0}^{D-1}\alpha_{i}\ket{i}\,.
\end{eqnarray}
Here $\alpha_{i}\in\mathComplex$ are complex probability amplitudes,
$\VVec{\alpha}\in\mathComplex^{D}$, and the $\{\ket{i}\}$ is an
orthonormal basis of states obeying 
\begin{equation}
\braket{i}{k}=\delta_{ik}\,.
\end{equation}

@@

In Sec.~\ref{CQC1qubitBloch.sec}, we focus on the discrete geometric
issues raised by the properties \ref{enu:-orthonormal-basis} and
\ref{enu:probability-amplitude} given above for CQT. In Sec.~\ref{sec:Quantum-Circuit-Model},
we describe product and entangled $n$-qubit states used in circuit
model, and the unitary matrices used to transform states as addressed
by by the property \ref{enu:unitary-matrix}. In Sec.~\ref{sec:fuzzy},
we introduces the important issues of \ref{enu:measurement-framework}
and the foundations of quantum probability space. To facilitate the
transition to DQT carried out in later sections, we concern ourselves
first with the properties of the simplest possible abstract state
object in CQT, the single qubit state.

@@

\subsection{The System Described by a Two-dimensional Hilbert Space}

A state in a two-dimensional Hilbert space, also called a qubit, already
provides access to a wealth of geometric information and context.
We write the single qubit state as 
\begin{align}
\ket{\psi_{1}} & =\alpha_{0}\ket{0}+\alpha_{1}\ket{1} & \alpha_{0},\alpha_{1}\in\mathbb{C}\,.\label{A1.eq}
\end{align}

A convenience for probability calculations and a necessity for computing
relative state properties is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $\rme^{\rmi\phi}$, that is, 
\begin{equation}
\left(\alpha_{0},\alpha_{1}\right)\sim\left(\rme^{\rmi\phi}\alpha_{0},\rme^{\rmi\phi}\alpha_{1}\right)\,.\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration, which we can write
down as follows: let 
\begin{align}
\alpha_{0} & =x_{0}+\rmi y_{0}\,, & \alpha_{1} & =x_{1}+\rmi y_{1}\,.\label{A4.eq}
\end{align}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
x_{0}^{\ 2}+y_{0}^{\ 2}+x_{1}^{\ 2}+y_{1}^{\ 2}=1\ .\label{A5.eq}
\end{equation}

There is a family of 6 equivalence classes of quadratic maps that
take the remaining 3 degrees of freedom in \Eq{A5.eq} and reduce
them to 2 degrees of freedom by effectively removing $\rme^{\rmi\phi}$
(``fibering out by the circle $\Sphere{1}$''). The standard form
of this class of maps (``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\ .\nonumber 
\end{eqnarray}
These transformed coordinates obey 
\begin{equation}
\|\Vec{X}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Figure \ref{hopfFibers.fig}
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point\/} $(\theta,\phi)$ on the surface
$X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

%% (We try to list the sources of all computed figures)%% Sources for hopfFibers.fig:  dqc-geom-2012/00mathFigs/Hopf-Fiber.nb
\begin{figure}[htb]
\centerline{ \includegraphics[width=0.48\columnwidth]{hopfS2}\hspace{0.03in}
\includegraphics[width=0.48\columnwidth]{hopfFibers}} \centerline{\hspace{1in}
(a) \hfill{}(b) \hspace{1in}} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration as a family
of circles (the paths of $\rme^{\rmi\phi}$), each corresponding to
a single point on the sphere in (a). Points in (a) are color coded
corresponding to circles in (b), e.g., one pole contains the red elliptical
circle that would become an infinite-radius circle in a slightly different
projection, and the opposite pole corresponds to the large perfectly
round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{eqnarray}
\hspace*{-0.5cm}\left(x_{0},y_{0},x_{1},y_{1}\right) & = & \left(\cos\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\sin\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\ \right.\nonumber \\
 &  & \left.\cos\frac{\theta-\phi}{2}\sin\frac{\psi}{2},\ \sin\frac{\theta-\phi}{2}\sin\frac{\psi}{2}\right),\label{A8a.eq}
\end{eqnarray}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\pi$, with $0\leq\frac{\theta+\phi}{2}<2\pi$ and $0\leq\frac{\theta-\phi}{2}<2\pi$,
we see that 
\begin{eqnarray}
\left(X,Y,Z\right)=\left(\cos\phi\sin\psi,\,\sin\phi\sin\psi,\,\cos\psi\right)\ .\label{A8b.eq}
\end{eqnarray}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to 
\begin{equation}
\ket{\psi_{1}}=\rme^{\rmi\phi}\cos\frac{\psi}{2}\ket{0}+\sin\frac{\psi}{2}\ket{1}\ ,\label{A9.eq}
\end{equation}
and an irreducible state can be represented as a point on a sphere,
as shown in Figure \ref{blochSphere.fig}(a).

\begin{figure}[thb]
\begin{centering}
\includegraphics[width=0.47\columnwidth]{blochSphere} \hspace{0.01in}
\includegraphics[width=0.47\columnwidth]{blochSphereArc} 
\par\end{centering}
\centerline{\hspace{1in} (a) \hfill{}(b) \hspace{1in}} \caption{(a) The conventional Bloch sphere with a unique state represented
by the point at the red sphere. (b) The geodesic shortest-distance
arc connecting two one-qubit quantum states.}
\label{blochSphere.fig} 
\end{figure}

@@

@@

For one qubit, we could easily solve the problem of reducing the full
unit-norm space to its irreducible components $\Vec{X}=(X,Y,Z)$ characterizing
the Bloch sphere. We have just argued that essentially the same process
is possible for $n$-qubits: in the abstract argument, we simply identify
the family of coefficients $\{\alpha_{i}\}$ as being the same if
they differ only by an overall phase $\rme^{\rmi\phi}$. However,
in practice this is not a construction that is easy to realize in
a practical computation. We now outline an explicit algorithm for
accomplishing the reduction to the irreducible $n$-qubit state space
$\CP{D-1}$; this construction will turn out to be useful for the
validation of our discrete results to follow below.

@@

We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $\rme^{\rmi\phi}$ from the $n$-qubit states. If we take
a particular weighting of the elements of the density matrix $\rho$,
we can construct a \textit{unit vector\/} of real dimension $D^{2}$
with the form: 
\begin{equation}
\Vec{X}=(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,\label{nqCoords.eq}
\end{equation}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
This construction gives an explicit embedding of the $(D-1)$-dimensional
complex, or $(2D-2)=(2^{n+1}-2)$-dimensional real, object in a real
space of dimension $D^{2}=2^{2n}$. However, this is somewhat subtle
because the vector is of unit length, so technically the embedding
space is a sphere of dimension $D^{2}-1=2^{2n}-1$ embedded in $\R^{D^{2}}$;
the one-qubit irreducible states could be represented in a 4D embedding,
but the magnitude of every coordinate would be one; furthermore, the
object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two\/} coordinate patches, one covering one pole of
$\Sphere{2}$ with coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & 1+0i\\
\alpha_{1} & = & x_{1}+iy_{1}
\end{array}
\end{eqnarray}
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & x_{0}+iy_{0}\\
\alpha_{1} & = & 1+0i\ .
\end{array}
\end{eqnarray}

We finally see that the irreducible $n$-qubit state space $\CP{D-1}$
is described by $D$ projectively equivalent coordinates, one of which
can always be scaled out to leave $(D-1)$ actual (complex) degrees
of freedom. We must choose, in turn, $D$ different local sets of
complex variables defined by taking the value $\alpha_{k}=1$, with
$k=0,\ldots,D-1$, and allowing the remaining $D-1$ complex (or $2D-2$
real) variables to run free. No single set of coordinates will work,
since the submanifold including $\alpha_{k}=0$ is undefined and another
coordinate system must be chosen to cover that coordinate patch. This
is a standard feature of the topology of non-trivial manifolds such
as $\CP{D-1}$ (see any textbook on geometry \cite{Berger}).
\end{comment}

\section{Quantum Circuit Model\label{sec:Quantum-Circuit-Model}}

@@

\todo{Explain enough background knowledge to support the discussion
about contextuality in the end of Sec.~\ref{modalquantum}.}
